---
title: "Warming Restructures Age-Dependent Disease Dynamics in a Multi-Pathogen System"
output: html_document
date: "2026-01-27"
---
```{r}
library(ggplot2)
library(cowplot)
library(car)
library(ggpubr)
library(RColorBrewer)
library(dplyr)
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(dplyr)
library(tidyr)
library(lme4)
library(lmerTest)
library(nlme)
library(emmeans)
library(ggplot2)

```


```{r}
df_GRmetadata<-read.csv("C:\\Users\\set2be\\OneDrive - University of North Carolina at Chapel Hill\\Chapter 1\\Formatted manuscript materials\\metadata.csv") 


df_GRmetadata$inoc<-factor(df_GRmetadata$inoc,levels=c("m","c","r","rc"))

names(df_GRmetadata)[2] <- 'temp'
names(df_GRmetadata)[1]<-'replicate'

df_GRmetadata$replicate<-factor(df_GRmetadata$replicate,levels=c("1","2","3","4","5","6"))

#df_GRmetadata

# Create treatment_block 
df_GRmetadata <- df_GRmetadata %>%
  mutate(treatment_block = interaction(inoc, age_gp, temp, sep = "_"))

```

###############################################
Survival 
###############################################
```{r}
data <- df_GRmetadata %>%
  group_by( replicate,inoc, age_gp, temp,treatment_block) %>%
  summarise(
    surviving = n(),
  ) %>%
  ungroup()

print(data)
```
##modeling 
```{r}

data$n_inoculated<- ifelse(data$replicate == 6, 35, 25)
data$died <- data$n_inoculated - data$surviving


survived <- glmer(
  cbind(surviving, died) ~ temp * age_gp * inoc + (1 | replicate/treatment_block),
  data = data,
  family = binomial
)

#summary(survived)
Anova(survived, type = 2)


emm <- emmeans(survived, ~temp * inoc, type = "response")
emm

```


###############################################
Disease Prevalence 
###############################################

######################Colletotrichum
```{r}
disease.long<-df_GRmetadata%>%
pivot_longer(cols=c(CollD5,CollD7,CollD10,CollD12,CollD15,CollD20),
              names_to="date",
                values_to="collprevalence")

disease.coll<-disease.long%>%
  mutate(date=factor(date,levels=c("CollD5","CollD7","CollD10","CollD12","CollD15","CollD20")))

prevcoll<-disease.coll%>%
  dplyr::group_by(replicate,date,inoc,age_gp,temp,treatment_block)%>%
  dplyr::summarize(prev=mean(collprevalence,na.rm=T))

prevcoll$date <- as.numeric(sub("CollD", "", prevcoll$date))


# Convert date column to character and remove "RhizD" string
prevcoll$date <- as.character(prevcoll$date)
prevcoll$date <- as.numeric(sub("CollD", "", prevcoll$date))
prevcoll$age_gp<-as.factor(prevcoll$age_gp)
prevcoll$temp<-as.factor(prevcoll$temp)

prevcoll



df_GRmetadata %>%
  dplyr::filter(inoc %in% c("c", "rc")) %>% 
  dplyr::group_by(inoc) %>%
  dplyr::summarize(
    N_infected = sum(CollD20 == 1, na.rm = TRUE),
    N_total = sum(!is.na(CollD20)),
    .groups = "drop"
  )
```




#Colletotrichum lesion development D20 glmer + EMM ggplot
```{r}
df_GRmetadata <- df_GRmetadata %>%
  mutate(id = seq_len(nrow(df_GRmetadata)))

prevcoll3_counts <- df_GRmetadata %>%
  filter(inoc %in% c("c","rc")) %>%
  group_by(temp, age_gp, inoc, replicate, id, treatment_block) %>%
  summarize(
    infected = sum(CollD20, na.rm = TRUE),
    total = sum(!is.na(CollD20)),
    .groups = "drop"
  )


prevcoll3_counts$age_gp <- factor(prevcoll3_counts$age_gp, levels = c("y","o"))
prevcoll3_counts$temp   <- factor(prevcoll3_counts$temp, levels = c("21","29"))
prevcoll3_counts$inoc   <- factor(prevcoll3_counts$inoc, levels = c("c","rc"))



glmm <- glmer(cbind(infected, total - infected) ~ temp * age_gp + temp*age_gp + temp + (1|replicate/treatment_block),
              data = prevcoll3_counts,
              family = binomial(link = "logit"))

summary(glmm)
Anova(glmm, type = 2)

emm <- emmeans(glmm, ~temp * age_gp, type = "response")


emm_df <- as.data.frame(emm) %>%
  mutate(
    pred_pct = prob,
    CI_lower = asymp.LCL,
    CI_upper = asymp.UCL
  )
emm_df


emm_df <- emm_df %>%
  mutate(
    Temperature_f = factor(
      temp,
      levels = c("21", "29"),
      labels = c("21°C", "29°C")
    ),
    Treatment_f = factor(
      age_gp,
      levels = c("y", "o"),
      labels = c(
        "Young",
        "Old"
      )
    )
  )


collemmplot<-ggplot(
  emm_df,
  aes(
    x = Temperature_f,
    y = pred_pct,
    group = Treatment_f,
    shape = Treatment_f
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.15,
    linewidth = 0.8
  ) +
  labs(
    title = expression(
      italic("C. cereale")),
    x = " ",
    y = " ",
    color = "Treatment",
    shape = "Treatment"
  ) +
  scale_color_manual(
    labels = c("Young", "Old")
  ) +
  scale_shape_manual(
    values = c(16, 17),
    labels = c("Young", "Old")
  ) +
  theme_classic(base_size = 14)+
  coord_cartesian(ylim = c(0, 1))

collemmplot




### inoc << inoc * temp significant, overlap in CI 

#glmm <- glmer(cbind(infected, total - infected) ~ temp * age_gp * inoc + (1|replicate/treatment_block),
 #             data = prevcoll3_counts,
#              family = binomial(link = "logit"))

#summary(glmm)
#Anova(glmm, type = 2)

#emm2 <- emmeans(glmm, ~temp * inoc, type = "response")


#emm_df2 <- as.data.frame(emm2) %>%
#  mutate(
#    pred_pct = prob,
#    CI_lower = asymp.LCL,
#    CI_upper = asymp.UCL
#  )
#emm_df2



```

#######################Rhizoctonia
```{r}
disease.long<-df_GRmetadata%>%
pivot_longer(cols=c(RhizD5,RhizD7,RhizD10,RhizD12,RhizD15,RhizD20),
              names_to="date",
                values_to="rhizprevalence")

disease.rhiz<-disease.long%>%
  mutate(date=factor(date,levels=c("RhizD5","RhizD7","RhizD10","RhizD12","RhizD15","RhizD20")))

prevrhiz<-disease.rhiz%>%
  dplyr::group_by(replicate,date,inoc,age_gp,temp,treatment_block)%>%
  dplyr::summarize(prev=mean(rhizprevalence,na.rm=T))

# Convert date column to character and remove "RhizD" string
prevrhiz$date <- as.character(prevrhiz$date)
prevrhiz$date <- as.numeric(sub("RhizD", "", prevrhiz$date))
prevrhiz$age_gp<-as.factor(prevrhiz$age_gp)
prevrhiz$temp<-as.factor(prevrhiz$temp)

prevrhiz


#sample size


df_GRmetadata %>%
  dplyr::filter(inoc %in% c("r", "rc")) %>% 
  dplyr::group_by(inoc) %>%
  dplyr::summarize(
    N_infected = sum(RhizD20 == 1, na.rm = TRUE),
    N_total = sum(!is.na(RhizD20)),
    .groups = "drop"
  )
```

#Rhizoctonia lesion development D20 glmer + EMM df + ggplot
```{r}
df_GRmetadata <- df_GRmetadata %>%
  mutate(id = seq_len(nrow(df_GRmetadata)))

prevrhiz3_counts <- df_GRmetadata %>%
  filter(inoc %in% c("r","rc")) %>%
  group_by(temp, age_gp, inoc, replicate, id,treatment_block) %>%
  summarize(
    infected = sum(RhizD20, na.rm = TRUE),
    total = sum(!is.na(RhizD20)),
    .groups = "drop"
  )

glmm <- glmer(cbind(infected, total - infected) ~ temp * age_gp  + (1|replicate/treatment_block),
              data = prevrhiz3_counts,
              family = binomial(link = "logit"))

summary(glmm)
Anova(glmm, type = 2)

emm <- emmeans(glmm, ~ temp * age_gp, type = "response")
#emm


emm_df <- as.data.frame(emm) %>%
  mutate(
    pred_pct = prob,
    CI_lower = asymp.LCL,
    CI_upper = asymp.UCL
  )
emm_df


emm_df <- emm_df %>%
  mutate(
    Temperature_f = factor(
      temp,
      levels = c("21", "29"),
      labels = c("21°C", "29°C")
    ),
    Treatment_f = factor(
      age_gp,
      levels = c("y", "o"),
      labels = c(
        "Young",
        "Old"
      )
    )
  )



rhizemmplot<-ggplot(
  emm_df,
  aes(
    x = Temperature_f,
    y = pred_pct,
    group = Treatment_f,
    shape = Treatment_f
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = CI_lower, ymax = CI_upper),
    width = 0.15,
    linewidth = 0.8
  ) +
  labs(
    title = expression(
      italic("R. solani")),
    x = " ",
    y = " ",
    color = "Treatment",
    shape = "Treatment"
  ) +
  scale_color_manual(
    labels = c("Young", "Old")
  ) +
  scale_shape_manual(
    values = c(16, 17),
    labels = c("Young", "Old")
  ) +
  theme_classic(base_size = 14)+
  coord_cartesian(ylim = c(0, 1))

rhizemmplot

```





#putting Coll and Rhiz EMM plots together
```{r}

prevemmfig <- ggarrange(
  collemmplot,
  rhizemmplot,
  nrow = 1, ncol = 2,
  common.legend = TRUE,
  legend = "right",
  labels = c("A.", "B."),
  font.label = list(size = 20, color = "black", face = "bold")
)

prevemmfig <- annotate_figure(
  prevemmfig,
  top = text_grob("",
                  face = "bold", size = 18),
  bottom = text_grob("Temperature Treatment", size = 14),
  left = text_grob("Proportion Infected/Inoculated Plants", size = 14, rot = 90)

)


prevemmfig

#ggsave("Figure2.png",plot=prevemmfig, width=6,height=4, dpi=600,bg = "white")

```


###############################################
AUDPS
###############################################
#making audps values
```{r}
# Select relevant columns
pla_data <- df_GRmetadata %>%
  select(replicate,inoc, age_gp, temp, plaD5, plaD7, plaD10, plaD12, plaD15, plaD20,dry.biomass,treatment_block)

#insert audps column 
audps_data <- pla_data %>%
  select(replicate, inoc, age_gp, temp, plaD5, plaD7, plaD10, plaD12, plaD15, plaD20,dry.biomass,treatment_block) %>%
  mutate(
    diff_plaD5 = abs(plaD5 - plaD5),
    diff_plaD7 = abs(plaD7 - plaD5),
    diff_plaD10 = abs(plaD10 - plaD5),
    diff_plaD12 = abs(plaD12 - plaD5),
    diff_plaD15 = abs(plaD15 - plaD5),
    diff_plaD20 = abs(plaD20 - plaD5),
    audps = diff_plaD5 + diff_plaD7 + diff_plaD10 + diff_plaD12 + diff_plaD15 + diff_plaD20,
     log_audps =log(audps + 1),
    log_dry_biomass = log(dry.biomass+1))

audps_data2<-subset(audps_data,audps!="NA")
audps_data2$replicate<-as.factor(audps_data2$replicate)
audps_data2$temp<-as.factor(audps_data2$temp)
audps_data2$age_gp<-as.factor(audps_data2$age_gp)

audps_infected <- audps_data2 %>%
  filter(audps > 0, inoc != "m")

nrow(audps_infected)

```
##disease-divided AUDPS 
```{r}
#Colletotrichum single infections
audps_coll <- audps_data %>%
  filter(audps > 0, inoc %in% c("c"))

head(audps_coll)

collaudpsmod<- lme(
  log_audps ~ temp * age_gp,
  random = ~1|replicate/treatment_block, 
  data= audps_coll,
  #weights = varIdent(form= ~1|inoc),
  method = "REML"
)

summary(collaudpsmod)
anova(collaudpsmod)


#Rhizoctonia single infections 
audps_rhiz<-audps_data%>%
  filter(audps>0,inoc %in% c("r"))

rhizaudpsmod<- lme(
  log_audps ~ temp * age_gp,
  random = ~1|replicate/treatment_block, 
  data= audps_rhiz,
  #weights = varIdent(form= ~1|inoc),
  method = "REML"
)

summary(rhizaudpsmod)
anova(rhizaudpsmod)


#Co-infections 
audps_co<-audps_data%>%
  filter(audps>0,inoc %in% c("rc"))

coaudps<- lme(
  log_audps ~ temp * age_gp,
  random = ~1|replicate/treatment_block, 
  data= audps_co,
 # weights = varIdent(form= ~1|inoc),
  method = "REML"
)

summary(coaudps)
anova(coaudps)
```
#Estimated Marginal Means
```{r}

#Coll 
#collemmtable<-emmeans(collaudpsmod,~temp|age_gp,type="response")
#collemmtable

#Rhiz
rhizemmtable<-emmeans(rhizaudpsmod,~temp|age_gp,type="response")
rhizemmtable

#back-transform from log scale
df<-as.data.frame(rhizemmtable)
rhizemmtable_trans<-df%>%
  mutate(
    emmean2 = exp(emmean),
    SE2 = exp(SE),
    lower = exp(lower.CL),
    upper = exp(upper.CL)
  )
rhizemmtable_trans


#Co-infections
coemmtable<-emmeans(coaudps,~temp|age_gp)
#coemmtable

#back-transform from log scale
df<-as.data.frame(coemmtable)
coemmtable_trans<-df%>%
  mutate(
    emmean2 = exp(emmean),
    SE2 = exp(SE),
    lower = exp(lower.CL),
    upper = exp(upper.CL)
  )
coemmtable_trans
```
#disease-divided AUDPS figure
```{r}


### --- GLOBAL COLORS & LABELS --- ###

infection_colors <- c(
  "c"  = "red",
  "r"  = "#009E73",
  "rc" = "#56B4E9"
)

infection_labels <- c(
  "c"  = expression(italic("C. cereale")),
  "r"  = expression(italic("R. solani")),
  "rc" = "Co-inoculation"
)

# Themes to remove axes
remove_x <- theme(
  axis.title.x = element_blank(),
  axis.text.x  = element_blank(),
  axis.ticks.x = element_blank()
)

remove_y <- theme(
  axis.title.y = element_blank(),
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank()
)



temp_labels <- c(
  "21" = expression(21 * degree * C),
  "29" = expression(29 * degree * C)
)

age_labels <-c(
  "o" = "Old",
  "y" = "Young"
)

### --- FIGURE 1: C. cereale --- ###

audps_coll$temp <- factor(audps_coll$temp, levels = c("21", "29"))

coll_audps_fig <- ggplot(audps_coll, aes(x = temp, y = audps, color = inoc)) +
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.3,
                                              dodge.width = 0.75), alpha = 0.3) +
  facet_grid(~age_gp, labeller = labeller(age_gp = age_labels)) +
  labs(x = NULL, y = NULL, color = "Infection") +
  ggtitle(expression(italic("C. cereale"))) +
  theme_classic() +
  ylim(0, 50) +
  scale_x_discrete(labels = c("21", "29")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +scale_color_manual(
  values = infection_colors,
  labels = infection_labels,
  limits = c("c", "r", "rc"),
  drop = FALSE
)

coll_audps_fig

### --- FIGURE 2: R. solani --- ###
### IMPORTANT: KEEP x-axis text here so "old" and "young" appear ###
audps_rhiz$temp <- factor(audps_rhiz$temp, levels = c("21", "29"))


rhiz_audps_fig <- ggplot(audps_rhiz, aes(x = temp, y = audps, color = inoc)) +
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.3, 
                                              dodge.width = 0.75), alpha = 0.3) +
  facet_grid(~age_gp, labeller = labeller(age_gp = age_labels)) +
  labs(x = NULL, y = NULL, color = "Infection") +
  ggtitle(expression(italic("R. solani"))) +
  theme_classic() +
    scale_x_discrete(labels = c("21", "29")) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)   # ← centers the title
  ) +
scale_color_manual(
  values = infection_colors,
  labels = infection_labels,
  limits = c("c", "r", "rc"),
  drop = FALSE
)

rhiz_audps_fig

### --- FIGURE 3: Co-infection --- ###
audps_co$temp <- factor(audps_co$temp, levels = c("21", "29"))

co_audps_fig <- ggplot(audps_co, aes(x = temp, y = audps, color = inoc)) +
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.3, 
                                              dodge.width = 0.75), alpha = 0.3) +
  facet_grid(~age_gp, labeller = labeller(age_gp = age_labels)) +
  labs(x = NULL, y = NULL, color = "Infection") +
  ggtitle("Co-infection") +
  theme_classic() +
   scale_x_discrete(labels = c("21", "29")) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)   # ← centers the title
  ) +
scale_color_manual(
  values = infection_colors,
  labels = infection_labels,
  limits = c("c", "r", "rc"),
  drop = FALSE
)

co_audps_fig

###### making the legend play nice
legend_df <- data.frame(
  temp = factor("21", levels = c("21","29")),
  audps = 1,
  inoc = factor(c("c","r","rc"), levels = c("c","r","rc"))
)

legend_plot <- ggplot(legend_df, aes(temp, audps, color = inoc)) +
  geom_point() +
  scale_color_manual(
    values = infection_colors,
    labels = infection_labels,
    limits = c("c","r","rc"),
    drop = FALSE
  ) +
  theme_void() +
  labs(color = "Inoculation Treatment")



### --- COMBINE THE PLOTS --- ###

audpsfig <- ggarrange(
  coll_audps_fig,
  rhiz_audps_fig,
  co_audps_fig,
  nrow = 1, ncol = 3,
  common.legend = TRUE,
legend = "right",
  legend.grob = get_legend(legend_plot),  labels = c("A.", "B.", "C."),
  font.label = list(size = 20, color = "black", face = "bold")
)


### --- ADD SHARED X AND Y AXES --- ###

audpsfig <- annotate_figure(
  audpsfig,
  top = text_grob("",
                  face = "bold", size = 18),
  bottom = text_grob("Temperature Treatment, °C", size = 16),
  left = text_grob("Disease Severity (AUDPS)", rot = 90, size = 16)
)

audpsfig


#ggsave("Figure3.png", audpsfig, width = 12, height = 5, dpi = 600, bg = "white")
```



###############################################
Dry Biomass
###############################################
```{r}

df_GRmetadata_dry.biomass <- df_GRmetadata %>%
 filter(!is.na(dry.biomass))%>%
  dplyr::ungroup()


df_GRmetadata_dry.biomass$age_gp<-factor(df_GRmetadata_dry.biomass$age_gp,levels=c("o","y"))
df_GRmetadata_dry.biomass$temp<-factor(df_GRmetadata_dry.biomass$temp,levels=c("29","21"))

df_GRmetadata_dry.biomass$temp <- factor(df_GRmetadata_dry.biomass$temp, levels = c("21", "29"))
df_GRmetadata_dry.biomass$age_gp <- factor(df_GRmetadata_dry.biomass$age_gp, levels = c("y", "o"))


df_GRmetadata_dry.biomass$temp <- as.factor(df_GRmetadata_dry.biomass$temp)
df_GRmetadata_dry.biomass$age_gp <- as.factor(df_GRmetadata_dry.biomass$age_gp)
df_GRmetadata_dry.biomass$inoc <- as.factor(df_GRmetadata_dry.biomass$inoc)
df_GRmetadata_dry.biomass$replicate<-as.factor(df_GRmetadata_dry.biomass$replicate)

df_GRmetadata_dry.biomass$logbiomass<-log(df_GRmetadata_dry.biomass$dry.biomass+ 1)
df_GRmetadata_dry.biomass <- df_GRmetadata_dry.biomass %>%
 filter(!is.na(logbiomass))
```
#plot
```{r}

#age-specific boxplots 
# Create separate plots for each age group

# Split the data into two separate data frames for each age group
df_GRmetadata_dry.biomass_o <- df_GRmetadata_dry.biomass %>% filter(age_gp == "o")
df_GRmetadata_dry.biomass_y <- df_GRmetadata_dry.biomass %>% filter(age_gp == "y")


# Plot for age_gp "o"
biomassold<-ggplot(df_GRmetadata_dry.biomass_o, aes(x = temp, y = dry.biomass, color = inoc)) + 
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.75), alpha = 0.3) +
  theme_classic() + 
  labs(x = " ", y = " ", color = "Inoculum") + 
  ggtitle("Old")+
  scale_y_continuous(limits = c(0, 10))+
     scale_x_discrete(labels = temp_labels) +
   scale_color_manual(values = c("c" = "red", "r" = "#009E73", "rc" = "#56B4E9", "m" = "#F0E442"),
labels = c(
    "m"  = "Mock",
    "r"  = expression(italic("R. solani")),
    "c"  = expression(italic("C. cereale")),
    "rc" = "Co-inoculated"
  ),
  limits = c("m","r","c","rc"),
  drop = FALSE
)
  
biomassold

# 1-10 scaled plot for age_gp "y" -> direct comparison
biomassyoung<-ggplot(df_GRmetadata_dry.biomass_y, aes(x = temp, y = dry.biomass, color = inoc)) + 
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.75), alpha = 0.3) +
  theme_classic() + 
  labs(x = " ", y = " ", color = "Inoculum") + 
  ggtitle("Young")+ 
  scale_y_continuous(limits = c(0, 10))+
     scale_x_discrete(labels = temp_labels) +
   scale_color_manual(values = c("c" = "red", "r" = "#009E73", "rc" = "#56B4E9", "m" = "#F0E442"),
labels = c(
    "m"  = "Mock",
    "r"  = expression(italic("R. solani")),
    "c"  = expression(italic("C. cereale")),
    "rc" = "Co-inoculated"
  ),
  limits = c("m","r","c","rc"),
  drop = FALSE
)


# Put figures together
biomass <- ggarrange(biomassold, biomassyoung, 
                      nrow = 1, ncol = 2, 
                      common.legend = TRUE, legend = "right",
                      labels = c("A.", "B."), 
                      font.label = list(size = 22, color = "black", face = "bold"))

biomass

biomass <- annotate_figure(
  biomass,
  top = text_grob(" ",
                  face = "bold", size = 18),
  bottom = text_grob("Temperature Treatment", size = 14),
  left = text_grob("Dry Biomass (g)", rot = 90, size = 14)
)

biomass

#ggsave("Figure4.png", biomass, width = 10, height = 6, dpi = 600, bg = "white")

```
#model
```{r}
mbiomass3<- lme(
  logbiomass ~ temp * age_gp * inoc,
  random = ~1 | replicate/treatment_block, 
  data = df_GRmetadata_dry.biomass,
  method = "REML",
)

summary(mbiomass3)
anova(mbiomass3)
```
#Estimated Marginal Means
```{r}
emm_all2 <- emmeans(mbiomass3, pairwise ~ temp|age_gp)
# View means
#summary(emm_all2)

##back-transform from log-scale:
emm_df <- summary(emm_all2$emmeans)

emm_df_bt <- emm_df %>%
  mutate(
    emmean_bt = exp(emmean),
    lower.CL_bt = exp(lower.CL),
    upper.CL_bt = exp(upper.CL)
  )

emm_df_bt
```

